```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(performance)
library(lme4)
library(partR2)
library(fitdistrplus)
library(glmmTMB)
# library(glmm.hp)
# library(r2glmm)
library(plotrix)
library(DHARMa)
library(MASS)

source("k_fold_cross_validation.R")
```

```{r}
natstories_rts <- read.csv("../naturalstories/naturalstories_RTS/processed_RTs.tsv", sep = "\t", header = TRUE) %>%
  distinct() %>%
  mutate(word_length = nchar(word))
```

```{r}
natstories_rts_by_zone <- natstories_rts %>%
  summarise(
    RT_mean = mean(RT),
    RT_sd = sd(RT),
    lower = RT_mean - RT_sd,
    upper = RT_mean + RT_sd,
    .by = c(item, zone)
  ) %>%
  mutate(zone_s = scale(zone))
```

```{r}
ggplot(natstories_rts_by_zone, aes(y = RT_mean, x = zone)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.4) +
  geom_line(aes(y = RT_mean), color = "grey", linewidth = .5) +
  facet_wrap(~item, scales = "free_x", ncol = 5) +
  coord_cartesian(ylim = c(100, 900)) +
  labs(
    title = "RTs by position in story",
    x = "Word Position",
    y = "Reading Time"
  ) + 
  theme_bw()
```

Remove outliers as in Oh and Schuler (2023)

```{r}
natstories_rts_clean <- natstories_rts %>%
  filter(RT > 200 & RT < 2000) %>%
  filter(correct >= 4)
```

Fit regression model on zone only

```{r}
natstories_rts_by_zone_clean <- natstories_rts_clean %>%
  summarise(
    RT_mean = mean(RT),
    RT_sd = sd(RT),
    lower = RT_mean - RT_sd,
    upper = RT_mean + RT_sd,
    .by = c(item, zone)
  ) %>%
  mutate(zone_s = scale(zone))
zone_model = lmer(RT_mean ~ zone + (1 + zone |item), data = natstories_rts_by_zone_clean)
```

Predict reading times from regression model

```{r}
grid_data <- natstories_rts_by_zone_clean %>% dplyr::select(item, zone)
RTs_pred = predict(zone_model, newdata = grid_data)
```

Calculate effective slopes (estimate of zone plus random slope per item)

```{r}
slope_zone <- fixef(zone_model)["zone"]
slopes_by_item <- ranef(zone_model)$item["zone"] + slope_zone
label_data = data.frame(
  item = seq(1,10),
  slope = round(slopes_by_item$zone,3)
) %>%
  mutate(
    label = str_glue("β = {slope}")
  )
```

```{r}
ggplot(natstories_rts_by_zone_clean, aes(y = RT_mean, x = zone)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.4) +
  geom_line(aes(y = RT_mean), color = "grey", size = .5) +
  geom_line(aes(x = zone, y = RTs_pred), color = "blue", size = 0.5) +
  geom_text(data = label_data, aes(x = 500, y = 800, label = label),
            color = "grey50", fontface = "bold", size = 2.5) +
  facet_wrap(~item, scales = "free_x", ncol = 5) +
  coord_cartesian(ylim = c(100, 1000)) +
  labs(
    title = "RTs by position in story",
    x = "Word Position",
    y = "Reading Time"
  ) + 
  theme_bw()
```

```{r}
natstories_model_surp <- bind_rows(
  read.csv("../annotated/surprisal/naturalstories/pythia-70m-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-70m"),
  read.csv("../annotated/surprisal/naturalstories/pythia-1.4b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-1.4b"),
  read.csv("../annotated/surprisal/naturalstories/pythia-12b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-12b"),
  read.csv("../annotated/surprisal/naturalstories/mgpt-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "mgpt-1.3B"),
  read.csv("../annotated/surprisal/naturalstories/mgpt-13B-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "mgpt-13B")
) %>%
  subset(select = -c(word)) %>%
  mutate(model = factor(model, levels = c("pythia-70m", "pythia-1.4b", "pythia-12b", "mgpt-1.3B", "mgpt-13B")))
```

```{r}
natstories_rts0 <- natstories_rts_clean %>%
  group_by(item, zone) %>%
  subset(select = -c(WorkerId, WorkTimeInSeconds)) %>%
  summarise(
    across(where(is.numeric), \(x) mean(x, rm.na = TRUE)),
    .groups  = "drop_last"
  )
natstories_plot_linear <- merge(x=natstories_rts0, y=natstories_model_surp, by = c("item", "zone"), all.x = TRUE)
```

```{r}
natstories_plot_linear %>%
  ungroup() %>%
  mutate(surprisal_bins = cut_number(surprisal, 25)) %>%
  group_by(
    surprisal_bins, model, item
    ) %>%
  mutate(mean_rt = mean(RT),
         mean_srp = mean(surprisal),
         se=std.error(RT),
         upper = mean_rt+1.96*se,
         lower = mean_rt-1.96*se) %>%
  ungroup() %>%
  ggplot(aes(x = mean_srp, y = mean_rt, ymin = lower, ymax =upper)) +
  stat_smooth(color="red", method="lm") +
  geom_errorbar(color="blue") + 
  geom_point(color="black", size = 0.5) +
  theme_bw() + 
  ylab("Reading Time (ms)") +
  xlab("Surprisal (bits)") +
  facet_grid(
    model ~ item,
    labeller = labeller(.multi_line = FALSE)
  )
```

```{r}
natstories_fit_df <- merge(x=natstories_rts_clean, y=natstories_model_surp %>% filter(model == "pythia-70m"), by = c("item", "zone"), all.x = TRUE) %>%
    group_by(item, sentence_id) %>%
    mutate(
      sentence_length = length(zone)
    ) %>%
  subset(select = -c(word, model, zone, WorkTimeInSeconds, correct)) %>%
  group_by(item, sentence_id, WorkerId) %>%
  summarise(
    # across(where(is.numeric), mean, na.rm  = TRUE), 
    sum_RT = sum(RT),
    sentence_length = mean(sentence_length),
    max_surprisal = max(surprisal),
    mean_surprisal = mean(surprisal),
    sum_surprisal = sum(surprisal),
    sentence_length_char = sum(word_length),
    sentence_length = mean(sentence_length),
    .groups = "drop_last"
  )
```

```{r}
# load and fix Souvik's naming conventions
natstories_ivs <- read.csv("../iv_temp_125_union/naturalstories/iv_temp_125.csv", sep = ",", header = TRUE) %>% 
  dplyr::select(!c(item_id, language, strategy, dataset, n_alternatives, n_complete_alternatives, n_incomplete_fallback_alternatives)) %>%
  dplyr::select(-contains("n_valid")) %>%
  mutate(item = as.integer(str_split_i(document_id, "_", 2))) %>%
  dplyr::select(!document_id) %>%
  rename(sentence_id = sentence_index)

# merge and scale
natstories_fit_df <- merge(x=natstories_fit_df, y=natstories_ivs, by = c("item", "sentence_id"), all.x = TRUE) %>%
  mutate(across(sum_RT:d_semantic_euclidean_mean, \(x) as.vector(scale(x)), .names = "{.col}_s"))
```


```{r}
ggplot(natstories_fit_df, aes(x=sum_RT)) +
  geom_histogram()
ggplot(natstories_fit_df, aes(x=log(sum_RT))) +
  geom_histogram() #+
  # scale_x_continuous(limits = c(100, 2000), breaks = seq(0, 2000, 50))
```

```{r}
descdist(natstories_fit_df$sum_RT, discrete=FALSE, boot=500)
descdist(log(natstories_fit_df$sum_RT), discrete=FALSE, boot=500)
```

```{r}
expl_df <- natstories_fit_df %>%
    filter(item %in% c(1,3,5,7,9))
```

```{r}
m0 <- lmer(
  sum_RT ~ 1 + (1|WorkerId) + (1|item),
  data = expl_df
  )

m_base <- lmer(
  sum_RT ~ sentence_length_s + (1|WorkerId) + (1|item),
  data = expl_df
)

m_base_char <- lmer(
  sum_RT ~ sentence_length_char_s + (1|WorkerId) + (1|item),
  data = expl_df
)

m_surp <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = expl_df
)

m_surp_char <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_length_char_s + (1|WorkerId) + (1|item),
  data = expl_df
)

anova(m0, m_base)
anova(m_base, m_base_char)
anova(m_base, m_surp)
anova(m_base_char, m_surp_char)
anova(m_surp, m_surp_char)
r2(m_base)
r2(m_surp)
r2(m_surp_char)
```
-> sentence length in characters is a little bit better.

```{r}
m_surp_inter <- lmer(
  sum_RT ~ sum_surprisal_s * sentence_length_char_s + (1|WorkerId) + (1|item),
  data = expl_df
)
anova(m_surp_inter, m_surp_char)
```
The interaction term only marginally improves model fit -> drop


```{r}
# check residuals
plot(simulateResiduals(m_surp_char))
```

-> Looks good!

```{r}
check_collinearity(m_surp_char)
```
High correlation, so maybe we need to use a combination of mean surprisal and sentence length

```{r}
m_surp_mean <- lmer(
  sum_RT ~ mean_surprisal_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = expl_df
)
m_surp_only <- lmer(
  sum_RT ~ sum_surprisal_s + (1|WorkerId) + (1|item),
  data = expl_df
)
anova(m_surp_only, m_surp_char)
r2(m_surp_only)
r2(m_surp_char)
```
This yields worse r2 and worse model fit, thus we will accept the collinearity and use both variables

```{r}
partR2(m_surp_char, data = expl_df,  partvars = c("sum_surprisal_s", "sentence_length_char_s"), nboot = 10)
```

```{r}
vec_ids <- unique(natstories_fit_df$WorkerId)
sample_id <- sample(vec_ids, 10)

natstories_fit_df %>%
  filter(
    WorkerId %in% sample_id,
    item == 1
    ) %>% 
  ggplot(aes(x = sum_surprisal_s, y = log(sum_RT))) +
  geom_point() +
  facet_wrap(~WorkerId)
```
No heteroscedasticity!


```{r}
m_ivs0 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d_semantic_cosine_mean_s + d1_recurrence_exponential_mean_s + d2_new_head_mean_s + (1|WorkerId) + (1|item), data = expl_df)
m_ivs1 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d_semantic_cosine_mean_s + d2_new_head_mean_s + (1|WorkerId) + (1|item), data = expl_df)
anova(m_ivs0, m_ivs1)
```
-> drop recurrence

```{r}
m_ivs0 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d_semantic_cosine_mean_s + d2_new_head_mean_s + (1|WorkerId) + (1|item), data = expl_df)
m_ivs1 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d2_new_head_mean_s + (1|WorkerId) + (1|item), data = expl_df)
anova(m_ivs0, m_ivs1)
```
--> keep cosine similarity of sent embeddings!

```{r}
m_ivs0 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d_semantic_cosine_mean_s + d2_new_head_mean_s + (1|WorkerId) + (1|item), data = expl_df)
m_ivs1 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d2_new_head_mean_s + d_semantic_cosine_mean_s + (1|WorkerId) + (1|item), data = expl_df)
anova(m_ivs0, m_ivs1)
```

```{r}
m_ivs0 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d_semantic_cosine_mean_s + d2_new_head_mean_s + (1|WorkerId) + (1|item), data = expl_df)
m_ivs1 <- lmer(sum_RT ~ sum_surprisal_s + sentence_length_char_s + d_lexical_1gram_mean_s + d_syntactic_1gram_mean_s + d_semantic_cosine_mean_s + d2_new_head_mean_s + d3_transition_mean_s +(1|WorkerId) + (1|item), data = expl_df)
anova(m_ivs0, m_ivs1)
summary(m_ivs1)
r2(m_ivs1)
```

```{r}
base_formula <- sum_RT ~ sum_surprisal_s + sentence_length_char_s + (1|WorkerId) + (1|item)
```

```{r}
model_formula_iv1 <- sum_RT ~ sum_surprisal_s + sentence_length_char_s + 
                              d_lexical_1gram_mean_s + 
                              d_syntactic_1gram_mean_s + 
                              d1_recurrence_exponential_mean_s + 
                              d_semantic_cosine_mean_s + 
                              d2_new_head_mean_s + 
                              d3_transition_mean_s + 
                              (1|WorkerId) + (1|item)
```


```{r}
results_iv1 <- model_cross_val_lmer(base_formula, model_formula_iv1, natstories_fit_df, group_var = "WorkerId", num_folds = 10)
cv_summary <- results_iv1$cv_summary
fixef_summary <- results_iv1$fixef_summary
cat("Mean delta log-likelihood per item (m1 - m0):", mean(cv_summary$delta_ll, na.rm = TRUE), "\n")
cat("Mean RMSE m0:", mean(cv_summary$rmse0, na.rm = TRUE), "\n")
cat("Mean RMSE m1:", mean(cv_summary$rmse1, na.rm = TRUE), "\n")
cat("Mean delta RMSE (m1 - m0):", mean(cv_summary$delta_rmse, na.rm = TRUE), "\n")
cat("Mean delta marginal R2:", mean(cv_summary$delta_r2m, na.rm = TRUE), "\n")
cat("Mean delta conditional R2:", mean(cv_summary$delta_r2c, na.rm = TRUE), "\n")
```
```{r}
fixef_summary %>%
  dplyr::select(!c("fold")) %>%
  group_by(term) %>%
  summarise(
    across(everything(), mean, na.rm = TRUE)
  ) %>%
  write.csv("results/naturalstories_fixefs.csv")

write.csv(cv_summary, "results/naturalstories_cv_summary.csv")
```


