```{r}
library(dplyr)
library(tidyverse)
library(performance)
library(lme4)
library(partR2)
library(fitdistrplus)
library(glmmTMB)
# library(glmm.hp)
# library(r2glmm)
library(plotrix)
library(DHARMa)
library(MASS)
```

```{r}
df_rts <- read.csv("../naturalstories/naturalstories_RTS/processed_RTs.tsv", sep = "\t", header = TRUE) %>%
  distinct() %>%
  mutate(word_length = nchar(word))
```

```{r}
df_rts_by_zone <- df_rts %>%
  summarise(
    RT_mean = mean(RT),
    RT_sd = sd(RT),
    lower = RT_mean - RT_sd,
    upper = RT_mean + RT_sd,
    .by = c(item, zone)
  ) %>%
  mutate(zone_s = scale(zone))
```

```{r}
ggplot(df_rts_by_zone, aes(y = RT_mean, x = zone)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.4) +
  geom_line(aes(y = RT_mean), color = "grey", size = .5) +
  facet_wrap(~item, scales = "free_x", ncol = 5) +
  coord_cartesian(ylim = c(100, 900)) +
  labs(
    title = "RTs by position in story",
    x = "Word Position",
    y = "Reading Time"
  ) + 
  theme_bw()
```

Remove outliers as in Oh and Schuler (2023)

```{r}
df_rts_clean <- df_rts %>%
  filter(RT > 200 & RT < 2000) %>%
  filter(correct >= 4)
```

Fit regression model on zone only

```{r}
df_rts_by_zone_clean <- df_rts_clean %>%
  summarise(
    RT_mean = mean(RT),
    RT_sd = sd(RT),
    lower = RT_mean - RT_sd,
    upper = RT_mean + RT_sd,
    .by = c(item, zone)
  ) %>%
  mutate(zone_s = scale(zone))
zone_model = lmer(RT_mean ~ zone + (1 + zone |item), data = df_rts_by_zone_clean)
```

Predict reading times from regression model

```{r}
grid_data <- df_rts_by_zone_clean %>% dplyr::select(item, zone)
RTs_pred = predict(zone_model, newdata = grid_data)
```

Calculate effective slopes (estimate of zone plus random slope per item)

```{r}
slope_zone <- fixef(zone_model)["zone"]
slopes_by_item <- ranef(zone_model)$item["zone"] + slope_zone
label_data = data.frame(
  item = seq(1,10),
  slope = round(slopes_by_item$zone,3)
) %>%
  mutate(
    label = str_glue("β = {slope}")
  )
```

```{r}
ggplot(df_rts_by_zone_clean, aes(y = RT_mean, x = zone)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.4) +
  geom_line(aes(y = RT_mean), color = "grey", size = .5) +
  geom_line(aes(x = zone, y = RTs_pred), color = "blue", size = 0.5) +
  geom_text(data = label_data, aes(x = 500, y = 800, label = label),
            color = "grey50", fontface = "bold", size = 2.5) +
  facet_wrap(~item, scales = "free_x", ncol = 5) +
  coord_cartesian(ylim = c(100, 1000)) +
  labs(
    title = "RTs by position in story",
    x = "Word Position",
    y = "Reading Time"
  ) + 
  theme_bw()
```

```{r}
df_surp_models <- bind_rows(
  read.csv("../annotated/surprisal/pythia-70m-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-70m"),
  read.csv("../annotated/surprisal/pythia-1.4b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-1.4b"),
  read.csv("../annotated/surprisal/pythia-12b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-12b"),
  read.csv("../annotated/surprisal/mgpt-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "mgpt-1.3B"),
  read.csv("../annotated/surprisal/mgpt-13B-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "mgpt-13B")
) %>%
  subset(select = -c(word)) %>%
  mutate(model = factor(model, levels = c("pythia-70m", "pythia-1.4b", "pythia-12b", "mgpt-1.3B", "mgpt-13B")))
```

```{r}
df_rts0 <- df_rts_clean %>%
  group_by(item, zone) %>%
  subset(select = -c(WorkerId, WorkTimeInSeconds)) %>%
  summarise(
    across(where(is.numeric), mean, rm.na = TRUE),
    .groups  = "drop_last"
  )

df_all0 <- merge(x=df_rts0, y=df_surp_models, by = c("item", "zone"), all.x = TRUE)
```

```{r}
df_all_plt <- df_all0 %>%
  ungroup() %>%
  mutate(surprisal_bins = cut_number(surprisal, 25)) %>%
  group_by(
    surprisal_bins, model, item
    ) %>%
  mutate(mean_rt = mean(RT),
         mean_srp = mean(surprisal),
         se=std.error(RT),
         upper = mean_rt+1.96*se,
         lower = mean_rt-1.96*se)
df_all_plt %>%
  ungroup() %>%
  ggplot(aes(x = mean_srp, y = mean_rt, ymin = lower, ymax =upper)) +
  stat_smooth(color="red", method="lm") +
  geom_errorbar(color="blue") + 
  geom_point(color="black", size = 0.5) +
  theme_bw() + 
  ylab("Reading Time (ms)") +
  xlab("Surprisal (bits)") +
  facet_grid(
    model ~ item,
    labeller = labeller(.multi_line = FALSE)
  )
```

```{r}
df_surp_pythia70m <- df_surp_models %>% filter(model == "pythia-70m") %>%
  group_by(item, sentence_id) %>%
  mutate(
    sentence_length = length(zone)
  )
df_all0 <- merge(x=df_rts_clean, y=df_surp_pythia70m, by = c("item", "zone"), all.x = TRUE)

df_all0_sent <- df_all0 %>%
  subset(select = -c(model, word, word_length, zone, WorkTimeInSeconds, correct)) %>%
  group_by(item, sentence_id, WorkerId) %>%
  summarise(
    # across(where(is.numeric), mean, na.rm  = TRUE), 
    sum_RT = sum(RT),
    sentence_length = mean(sentence_length),
    max_surprisal = max(surprisal),
    mean_surprisal = mean(surprisal),
    sum_surprisal = sum(surprisal),
    sentence_length = mean(sentence_length),
    .groups = "drop_last"
  )
```

```{r}
files <- list.files("../iv_ns/", pattern = ".csv", full.names = TRUE)
df_iv <- files %>%
  map_df(~read_csv(., show_col_types = FALSE), .id = "item")
df_iv <- df_iv %>%
  subset(select = -c(dataset, annotation_condition, group_id,
                                   numeric_group_id, context_id, story_id,
                                   target_id, target_variant, target_text,
                                   mean_acceptability, median_acceptability,
                                   n_alternatives, source_schema_version,
                                   metric_version, table_schema_version)) %>%
  rename(., sentence_id = sentence_index)
df_all0_sent <- merge(x=df_all0_sent, y=df_iv, by = c("item", "sentence_id"), all.x = TRUE)
```

```{r}

print(sum(is.na(df_all0_sent$sentence_id)))
print(sum(is.na(df_all0_sent$sentence_length)))

print(sd(df_all0_sent$sentence_id))
print(sd(df_all0_sent$sentence_length))

df_all0_sent <- df_all0_sent %>%
  mutate(
    sum_surprisal_s = as.vector(scale(sum_surprisal)),
    mean_surprisal_s = as.vector(scale(mean_surprisal)),
    max_surprisal_s = as.vector(scale(max_surprisal)),
    sentence_length_s = as.vector(scale(sentence_length)),
    sentence_id_s = as.vector(scale(sentence_id)),
    # surprisal_s = as.vector(scale(surprisal)),
    # word_length_s = as.vector(scale(word_length)),
    # sentence_position_s = as.vector(scale(sentence_position)),
    # zone_s= as.vector(scale(zone)),
    # log_freq_s = as.vector(scale(log_freq)),
    iv_recurrence_mean_s = as.vector(scale(iv_recurrence_mean)),
    iv_feature_role_form_mean_s = as.vector(scale(iv_feature_role_form_mean)),
    iv_feature_role_form_no_other_mean_s = as.vector(scale(iv_feature_role_form_no_other_mean)),
    iv_feature_role_only_mean_s = as.vector(scale(iv_feature_role_only_mean)),
    iv_new_entity_mean_s = as.vector(scale(iv_new_entity_mean)),
    iv_recurrence_p80_s = as.vector(scale(iv_recurrence_p80)),
    iv_feature_role_form_p80_s = as.vector(scale(iv_feature_role_form_p80)),
    iv_feature_role_no_other_p80_s = as.vector(scale(iv_feature_role_form_no_other_p80)),
    iv_feature_role_only_p80_s = as.vector(scale(iv_feature_role_only_p80)),
    iv_new_entity_p80_s = as.vector(scale(iv_new_entity_p80)),
  ) # %>%
  # mutate_if(is.numeric, ~replace_na(., 0.))
df_all0_sent$sentence_id_s <- as.vector(scale(df_all0_sent$sentence_id))
df_all0_sent$sentence_length_s <- as.vector(scale(df_all0_sent$sentence_length))
df_all0_sent <- df_all0_sent %>% mutate_if(is.numeric, ~replace_na(., 0.))
```

```{r}
ggplot(df_all0_sent, aes(x=log(sum_RT))) +
  geom_histogram() #+
  #scale_x_continuous(limits = c(100, 2000), breaks = seq(0, 2000, 50)) 
```

```{r}
descdist(df_all0_sent$sum_RT, discrete=FALSE, boot=500)
descdist(log(df_all0_sent$sum_RT), discrete=FALSE, boot=500)
```

```{r}
df_fit <- df_all0_sent %>%
    filter(item %in% c(1,3,5,7,9))
```

```{r}
base_model <- lmer(
  sum_RT ~ sentence_id_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = df_fit
  )

surp_model <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_id_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = df_fit
)

surp_model_max <- lmer(
  sum_RT ~ max_surprisal_s + sentence_id_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = df_fit
)

plot(simulateResiduals(surp_model))
plot(simulateResiduals(surp_model_max))
```

```{r}
check_collinearity(surp_model)
check_collinearity(surp_model_max)
```

```{r}
anova(base_model, surp_model)
# base_model: sum_RT ~ sentence_id_s + sentence_length_s + (1 | WorkerId) + (1 | item)
# surp_model: sum_RT ~ sum_surprisal_s + sentence_id_s + sentence_length_s + (1 | WorkerId) + (1 | item)
#            npar    AIC    BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# base_model    6 383625 383673 -191806    383613                         
# surp_model    7 379528 379584 -189757    379514 4099.1  1  < 2.2e-16 ***
# --> surprisal model improves over the base model by quite a bit

anova(surp_model, surp_model_max)
# surp_model: sum_RT ~ sum_surprisal_s + sentence_id_s + sentence_length_s + (1 | WorkerId) + (1 | item)
# surp_model_max: sum_RT ~ max_surprisal_s + sentence_id_s + sentence_length_s + (1 | WorkerId) + (1 | item)
#                npar    AIC    BIC  logLik -2*log(L) Chisq Df Pr(>Chisq)
# surp_model        7 379528 379584 -189757    379514                    
# surp_model_max    7 382702 382758 -191344    382688     0  0    
# --> max doesn't improve RT fit, use sum of surprisals
partR2(surp_model, data = df_fit,  partvars = c("sum_surprisal_s", "sentence_id_s", "sentence_length_s"), nboot = 10)
```

```{r}
surp_model_log <- glmmTMB(
  sum_RT ~ sum_surprisal_s + sentence_id_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = df_fit,
  family = Gamma(link = "log")
)
```

```{r}
simulateResiduals(surp_model_log, plot = TRUE)
```

```{r}
vec_ids <- unique(df_all0$WorkerId)
sample_id <- sample(vec_ids, 10)

df_all0_sent %>%
  filter(
    WorkerId %in% sample_id,
    item == 1
    ) %>% 
  ggplot(aes(x = sum_surprisal_s, y = log(sum_RT))) +
  geom_point() +
  facet_wrap(~WorkerId)
```

```{r}
surp_model_disp <- glmmTMB(
  sum_RT ~ sum_surprisal_s + sentence_id_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = df_fit, family = Gamma(link = "log"), 
  dispformula = ~ sum_surprisal_s
  )
res_gamma <- simulateResiduals(surp_model_disp, plot = TRUE)
```

Actually the lmer looks better, Cullen and Frey plots indicate that the sum of reading times shouldn't be log-transformed!

```{r}
anova(surp_model_disp,surp_model_log)
```

```{r}
surp_model <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_id_s #+ sentence_length_s 
  + (1|WorkerId)
  + (1|item),
  data = df_all0_sent
)

surp_model_mean <- lmer(
  sum_RT ~ mean_surprisal_s + sentence_id_s + sentence_length_s 
  + (1|WorkerId)
  + (1|item),
  data = df_all0_sent
)

check_collinearity(surp_model_mean)

iv_model_recurrence_mean <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_id_s #+ sentence_length_s 
  + iv_recurrence_mean_s
  + (1|WorkerId)
  + (1|item),
  data = df_all0_sent
)

iv_model_feature_role_only_mean <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_id_s #+ sentence_length_s 
  + iv_feature_role_only_mean_s
  + (1|WorkerId)
  + (1 |item),
  data = df_all0_sent
)

iv_model_role_comb <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_id_s #+ sentence_length_s 
  + iv_recurrence_mean_s + iv_feature_role_only_mean_s
  + (1|WorkerId)
  + (1 |item),
  data = df_all0_sent
)

# iv_model_new_entity_mean <- lmer(
#   sum_RT ~ mean_surprisal_s + sentence_id_s + sentence_length_s 
#   + iv_new_entity_mean_s
#   + (1|WorkerId)
#   + (1 |item),
#   data = df_all0
# )

# not scaled :(
# iv_model_rec_min <- lmer(
#   sum_RT ~ sum_surprisal_s + sentence_id_s + sentence_length_s 
#   + iv_recurrence_min_s
#   + (1|WorkerId) + (1|item),
#   data = df_fit
# )
```

Separating sentence length and surprisal helps a bit!

```{r}
anova(surp_model, iv_model_recurrence_mean)
summary(iv_model_recurrence_mean)

anova(surp_model, iv_model_feature_role_only_mean)
summary(iv_model_feature_role_only_mean)

# anova(surp_model_mean, iv_model_feature_role_only_mean)
# summary(iv_model_feature_role_form_mean)
# 
# anova(surp_model_mean, iv_model_new_entity_mean)
# summary(iv_model_new_entity_mean)
```

We conclude that using the 80 percentile recurrence yields a worse fit! Let's look at R2:

```{r}
partR2(surp_model, data = df_all0_sent,  partvars = c("sum_surprisal_s", "sentence_id_s"), nboot = 10)
partR2(iv_model_recurrence_mean, data = df_all0_sent,  partvars = c("sum_surprisal_s", "sentence_id_s", "iv_recurrence_mean_s"), nboot = 10)
```

```{r}
# df_fit0 <- df_fit %>%
#   group_by(item, zone, sentence_id, sentence_position, word) %>%
#   summarise(
#     across(
#       RT:iv_new_entity_p80_s, ~ mean(.x, na.rm = TRUE)
#     ),
#     .groups = "drop_last"
#   )
```

Averaging over participants (WorkerID) yields much better R2

```{r}
# base_model0_gau <- glmmTMB(
#   RT ~ sentence_position_s + zone_s + log_freq_s + word_length_s + (1|item),
#   data = df_fit0, family = gaussian
#   )
# base_model0_log <- glmmTMB(
#   RT ~ sentence_position_s + zone_s + log_freq_s + word_length_s + (1|item),
#   data = df_fit0, family = lognormal
#   )
# 
# surp_model0_gau <- glmmTMB(
#   RT ~ surprisal_s + sentence_position_s + zone_s + log_freq_s + word_length_s + (1|item),
#   data = df_fit0, family = gaussian
#   )
# surp_model0_log <- glmmTMB(
#   RT ~ surprisal_s + sentence_position_s + zone_s + log_freq_s + word_length_s + (1|item),
#   data = df_fit0, family = lognormal
#   )

# surp_model0_log <- glmmTMB(
#   RT ~ surprisal_s + sentence_position_s + zone_s + log_freq_s + word_length_s + (1|item),
#   data = df_fit0, family = Gamma(link = "log")
#   )

# res_gau <- simulateResiduals(surp_model0_log, plot = TRUE)
# res_log <- simulateResiduals(surp_model0_log_d, plot = TRUE)
```

```{r}
# anova(base_model0_log,surp_model0_log)
# fixef(surp_model0_log)
# r2(base_model0_log)
# r2(surp_model0_log)
#partR2(surp_model0, data = df_fit0,  partvars = c("surprisal", "sentence_position", "zone", "log_freq", "word_length"), nboot = 10)

# check if de varda and marelli included zone!
# --> They don't
```

```{r}
# surp_model0_s <- lmer(
#   RT ~ surprisal_s + sentence_position_s + zone_s + log_freq_s + word_length_s + (1|item),
#   data = df_fit0
#   )
# 
# feature_role_model0_s <- lmer(
#   RT ~ surprisal_s + word_length_s + sentence_position_s + log_freq_s + zone_s
#   + iv_feature_role_form_p80_s 
#   + (1|item),
#     data = df_fit0
#   )
# recurrence_model0_s <- lmer(
#   RT ~ surprisal_s + word_length_s + sentence_position_s + log_freq_s + zone_s
#   + iv_recurrence_p80_s 
#   + (1|item),
#     data = df_fit0
#   )
# new_entity_model0_s <- lmer(
#   RT ~ surprisal_s + word_length_s + sentence_position_s + log_freq_s + zone_s
#   + iv_new_entity_p80_s 
#   + (1|item),
#   data = df_fit0
#   )
```

Adding entity IVs marginally improves model fit, but r2 for iv features is very small

```{r}
# anova(feature_role_model0_s,surp_model0_s, test = "Chisq")
# fixef(feature_role_model0_s)
# partR2(feature_role_model0_s, data = df_fit0,  partvars = c("surprisal_s", "sentence_position_s", "log_freq_s", "zone_s", "word_length_s", "iv_feature_role_form_p80_s"), nboot = 10)

# anova(recurrence_model0_s,surp_model0_s, test = "Chisq")
# fixef(recurrence_model0_s)
# partR2(recurrence_model0_s, data = df_fit0,  partvars = c("surprisal_s", "sentence_position_s", "log_freq_s", "zone_s", "word_length_s", "iv_recurrence_p80_s"), nboot = 10)

# anova(new_entity_model0_s,surp_model0_s, test = "Chisq")
# fixef(new_entity_model0_s)
# partR2(new_entity_model0_s, data = df_fit0,  partvars = c("surprisal_s", "sentence_position_s", "log_freq_s", "zone_s", "word_length_s", "iv_new_entity_p80_s"), nboot = 10)
```

```{r}
story_ids <- unique(df_all$item)

cross_validate <- function(df) {
  
  df_res <- data.frame(
    test_id = numeric(),
    model = character(),
    rmse_base = numeric(),
    rmse_iv = numeric(),
    log_lik_base = numeric(),
    log_lik_iv = numeric(),
    r2_base = numeric(),
    r2_iv = numeric()
  )
  
  for (test_id in 1:10) {
    train_ids <- setdiff(story_ids, c(test_id))
    df_test <- df %>% filter(item == test_id)
    df_train <- df %>% filter(item %in% train_ids)
    
    # base model with only surprisal, position of the sentence in the story, 
    # and sentence length as predictors
    base_model <- lmer(
      sum_RT ~ mean_surprisal_s + sentence_id_s + sentence_length_s + (1|item),
      data = df_train, REML = FALSE
    )
    
    # evaluate base model
    log_lik_base <- logLik(base_model)
    pred_RTs_base <- predict(base_model, newdata = df_test, allow.new.levels = TRUE)
    rmse_base <- sqrt(mean((pred_RTs_base-df_test$sum_RT)^2))
    
    iv_model_recurrence_mean <- lmer(
      sum_RT ~ mean_surprisal_s + sentence_id_s + sentence_length_s + iv_recurrence_mean_s + (1|item),
      # + (1|WorkerId),
      data = df_train, REML = FALSE
    )
    
    iv_model_feature_role_form_mean <- lmer(
      sum_RT ~ mean_surprisal_s + sentence_id_s + sentence_length_s + iv_feature_role_form_mean_s + (1 |item),
      # + (1|WorkerId), 
      data = df_train, REML = FALSE
    )
    
    iv_model_new_entity_mean <- lmer(
      sum_RT ~ mean_surprisal_s + sentence_id_s + sentence_length_s + iv_new_entity_mean_s + (1 |item),
      # + (1|WorkerId), 
      data = df_train, REML = FALSE
    )
    
    model_list <- list(list("recurrence", iv_model_recurrence_mean), 
                       list("feature_role", iv_model_feature_role_form_mean), 
                       list("new_entity", iv_model_new_entity_mean))
    
    for (iv in model_list) {
      iv_name <- iv[[1]]
      iv_model <- iv[[2]]
      log_lik_iv <- logLik(iv_model)
      pred_RTs_iv <- predict(iv_model, newdata = df_test, allow.new.levels = TRUE)
      rmse_iv <- sqrt(mean((pred_RTs_iv-df_test$sum_RT)^2))
      df_res <- df_res %>%
        add_row(
          test_id = test_id,
          model = iv_name,
          rmse_base = rmse_base,
          rmse_iv = rmse_iv,
          log_lik_base = as.numeric(log_lik_base),
          log_lik_iv = as.numeric(log_lik_iv),
          r2_base = as.numeric(r2(base_model)),
          r2_iv = as.numeric(r2(iv_model))
        )
    }
  }
  return(df_res)
}
df_res <- cross_validate(df_all0)
```
