```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(performance)
library(lmerTest)
library(lme4)
library(partR2)
library(fitdistrplus)
library(glmmTMB)
# library(glmm.hp)
# library(r2glmm)
library(plotrix)
library(DHARMa)
library(MASS)
library(glue)

source("k_fold_cross_validation.R")
```

```{r}
natstories_rts <- read.csv("../naturalstories/naturalstories_RTS/processed_RTs.tsv", sep = "\t", header = TRUE) %>%
  distinct() %>%
  mutate(word_length = nchar(word))
```

```{r}
natstories_rts_by_zone <- natstories_rts %>%
  summarise(
    RT_mean = mean(RT),
    RT_sd = sd(RT),
    lower = RT_mean - RT_sd,
    upper = RT_mean + RT_sd,
    .by = c(item, zone)
  ) %>%
  mutate(zone_s = scale(zone))
```

```{r}
ggplot(natstories_rts_by_zone, aes(y = RT_mean, x = zone)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.4) +
  geom_line(aes(y = RT_mean), color = "grey", linewidth = .5) +
  facet_wrap(~item, scales = "free_x", ncol = 5) +
  coord_cartesian(ylim = c(100, 900)) +
  labs(
    title = "RTs by position in story",
    x = "Word Position",
    y = "Reading Time"
  ) + 
  theme_bw()
```

Remove outliers as in Oh and Schuler (2023)

```{r}
natstories_rts_clean <- natstories_rts %>%
  filter(RT > 200 & RT < 2000) %>%
  filter(correct >= 4)
```

Fit regression model on zone only

```{r}
natstories_rts_by_zone_clean <- natstories_rts_clean %>%
  summarise(
    RT_mean = mean(RT),
    RT_sd = sd(RT),
    lower = RT_mean - RT_sd,
    upper = RT_mean + RT_sd,
    .by = c(item, zone)
  ) %>%
  mutate(zone_s = scale(zone))
zone_model = lmer(RT_mean ~ zone + (1 + zone |item), data = natstories_rts_by_zone_clean)
```

Predict reading times from regression model

```{r}
grid_data <- natstories_rts_by_zone_clean %>% dplyr::select(item, zone)
RTs_pred = predict(zone_model, newdata = grid_data)
```

Calculate effective slopes (estimate of zone plus random slope per item)

```{r}
slope_zone <- fixef(zone_model)["zone"]
slopes_by_item <- ranef(zone_model)$item["zone"] + slope_zone
label_data = data.frame(
  item = seq(1,10),
  slope = round(slopes_by_item$zone,3)
) %>%
  mutate(
    label = str_glue("β = {slope}")
  )
```

```{r}
ggplot(natstories_rts_by_zone_clean, aes(y = RT_mean, x = zone)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.4) +
  geom_line(aes(y = RT_mean), color = "grey", size = .5) +
  geom_line(aes(x = zone, y = RTs_pred), color = "blue", size = 0.5) +
  geom_text(data = label_data, aes(x = 500, y = 800, label = label),
            color = "grey50", fontface = "bold", size = 2.5) +
  facet_wrap(~item, scales = "free_x", ncol = 5) +
  coord_cartesian(ylim = c(100, 1000)) +
  labs(
    title = "RTs by position in story",
    x = "Word Position",
    y = "Reading Time"
  ) + 
  theme_bw()
```

```{r}
natstories_model_surp <- bind_rows(
  read.csv("../annotated/surprisal/naturalstories/pythia-400m-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-400m", checkpoint = "step1000"),
  read.csv("../annotated/surprisal/naturalstories/pythia-400m-deduped-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-400m", checkpoint = "full model"),
  read.csv("../annotated/surprisal/naturalstories/pythia-1.4b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-1.4b", checkpoint = "step1000"),
  read.csv("../annotated/surprisal/naturalstories/pythia-1.4b-deduped-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-1.4b", checkpoint = "full model"),
  read.csv("../annotated/surprisal/naturalstories/pythia-6.9b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-6.9b", checkpoint = "step1000"),
  read.csv("../annotated/surprisal/naturalstories/pythia-6.9b-deduped-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-6.9b", checkpoint = "full model"),
  read.csv("../annotated/surprisal/naturalstories/pythia-12b-deduped-step1000-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-12b", checkpoint = "step1000"),
  read.csv("../annotated/surprisal/naturalstories/pythia-12b-deduped-full.tsv", sep = "\t", header = TRUE) %>%
    mutate(model = "pythia-12b", checkpoint = "full model"),
  # read.csv("../annotated/surprisal/naturalstories/mgpt-full.tsv", sep = "\t", header = TRUE) %>%
  #   mutate(model = "mgpt-1.3B"),
  # read.csv("../annotated/surprisal/naturalstories/mgpt-13B-full.tsv", sep = "\t", header = TRUE) %>%
  #   mutate(model = "mgpt-13B")
) %>%
  subset(select = -c(word)) %>%
  mutate(model = factor(model, levels = c("pythia-400m", "pythia-1.4b", "pythia-6.9b", "pythia-12b")))
```

```{r}
natstories_rts0 <- natstories_rts_clean %>%
  group_by(item, zone) %>%
  subset(select = -c(WorkerId, WorkTimeInSeconds)) %>%
  summarise(
    across(where(is.numeric), \(x) mean(x, rm.na = TRUE)),
    .groups  = "drop_last"
  )
natstories_plot_linear <- merge(x=natstories_rts0, y=natstories_model_surp, by = c("item", "zone"), all.x = TRUE)
```

```{r}
natstories_plot_linear %>%
  ungroup() %>%
  mutate(surprisal_bins = cut_number(surprisal, 25)) %>%
  group_by(
    surprisal_bins, model, item
    ) %>%
  mutate(mean_rt = mean(RT),
         mean_srp = mean(surprisal),
         se=std.error(RT),
         upper = mean_rt+1.96*se,
         lower = mean_rt-1.96*se) %>%
  ungroup() %>%
  ggplot(aes(x = mean_srp, y = mean_rt, ymin = lower, ymax =upper)) +
  stat_smooth(color="red", method="lm") +
  geom_errorbar(color="blue") + 
  geom_point(color="black", size = 0.5) +
  theme_bw() + 
  ylab("Reading Time (ms)") +
  xlab("Surprisal (bits)") +
  facet_grid(
    model ~ item,
    labeller = labeller(.multi_line = FALSE)
  )
```

```{r}
natstories_fit_df <- merge(x=natstories_rts_clean, y=natstories_model_surp, by = c("item", "zone"), all.x = TRUE) %>%
    group_by(item, sentence_id, model, checkpoint) %>%
    mutate(
      sentence_length = length(zone)
    ) %>%
  subset(select = -c(word, zone, WorkTimeInSeconds, correct)) %>%
  group_by(item, sentence_id, model, checkpoint, WorkerId) %>%
  summarise(
    # across(where(is.numeric), mean, na.rm  = TRUE), 
    sum_RT = sum(RT),
    sentence_length = mean(sentence_length),
    max_surprisal = max(surprisal),
    mean_surprisal = mean(surprisal),
    sum_surprisal = sum(surprisal),
    sentence_length_char = sum(word_length),
    sentence_length = mean(sentence_length),
    .groups = "drop_last"
  )
```

```{r}
# load and fix Souvik's naming conventions
natstories_ivs <- read.csv("../iv_values/naturalstories/iv_temp_125.csv", sep = ",", header = TRUE) %>% 
  dplyr::select(!c(item_id, language, strategy, dataset)) %>%
  dplyr::select(-contains("n_valid")) %>%
  mutate(item = as.integer(str_split_i(document_id, "_", 2))) %>%
  dplyr::select(!document_id) %>%
  rename(sentence_id = sentence_index)

# merge and scale
natstories_fit_df <- merge(x=natstories_fit_df, y=natstories_ivs, by = c("item", "sentence_id"), all.x = TRUE) %>%
  mutate(across(sum_RT:d_semantic_euclidean_mean, \(x) as.vector(scale(x)), .names = "{.col}_s"))
```


```{r}
ggplot(natstories_fit_df, aes(x=sum_RT)) +
  geom_histogram()
ggplot(natstories_fit_df, aes(x=log(sum_RT))) +
  geom_histogram() #+
  # scale_x_continuous(limits = c(100, 2000), breaks = seq(0, 2000, 50))
```

```{r}
descdist(natstories_fit_df$sum_RT, discrete=FALSE, boot=500)
descdist(log(natstories_fit_df$sum_RT), discrete=FALSE, boot=500)
```

```{r}
expl_df <- natstories_fit_df %>%
    filter(item %in% c(1,3,5,7,9), model == "pythia-12b")
```

```{r}
m0 <- lmer(
  sum_RT ~ 1 + (1|WorkerId) + (1|item),
  data = expl_df
  )

m_base <- lmer(
  sum_RT ~ sentence_length_s + (1|WorkerId) + (1|item),
  data = expl_df
)

m_base_char <- lmer(
  sum_RT ~ sentence_length_char_s + (1|WorkerId) + (1|item),
  data = expl_df
)

m_surp <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = expl_df
)

m_surp_char <- lmer(
  sum_RT ~ sum_surprisal_s + sentence_length_char_s + (1|WorkerId) + (1|item),
  data = expl_df
)

anova(m0, m_base)
anova(m_base, m_base_char)
anova(m_base, m_surp)
anova(m_base_char, m_surp_char)
anova(m_surp, m_surp_char)
r2(m_base)
r2(m_surp)
r2(m_surp_char)
```
-> sentence length in characters is a little bit better.

```{r}
m_surp_inter <- lmer(
  sum_RT ~ sum_surprisal_s * sentence_length_char_s + (1|WorkerId) + (1|item),
  data = expl_df
)
anova(m_surp_inter, m_surp_char)
```
The interaction term only marginally improves model fit -> drop


```{r}
# check residuals
plot(simulateResiduals(m_surp_char))
```

-> Looks good!

```{r}
check_collinearity(m_surp_char)
```
High correlation, so maybe we need to use a combination of mean surprisal and sentence length

```{r}
m_surp_mean <- lmer(
  sum_RT ~ mean_surprisal_s + sentence_length_s + (1|WorkerId) + (1|item),
  data = expl_df
)
m_surp_only <- lmer(
  sum_RT ~ sum_surprisal_s + (1|WorkerId) + (1|item),
  data = expl_df
)
anova(m_surp_only, m_surp_char)
r2(m_surp_only)
r2(m_surp_char)
```
This yields worse r2 and worse model fit, thus we will accept the collinearity and use both variables

```{r}
partR2(m_surp_char, data = expl_df,  partvars = c("sum_surprisal_s", "sentence_length_char_s"), nboot = 10)
```

```{r}
vec_ids <- unique(natstories_fit_df$WorkerId)
sample_id <- sample(vec_ids, 10)

natstories_fit_df %>%
  filter(
    WorkerId %in% sample_id,
    item == 1
    ) %>% 
  ggplot(aes(x = sum_surprisal_s, y = log(sum_RT))) +
  geom_point() +
  facet_wrap(~WorkerId)
```
No heteroscedasticity!

```{r}
iv_formula <- sum_RT ~ sum_surprisal_s + sentence_length_char_s + 
                              d1_recurrence_uniform_mean_s +
                              d1_recurrence_exponential_mean_s +
                              d2_new_head_mean_s +
                              d3_transition_mean_s +
                              d_lexical_1gram_mean_s +
                              d_lexical_2gram_mean_s +
                              d_lexical_3gram_mean_s +
                              d_pos_overlap_1gram_mean_s +
                              d_pos_overlap_2gram_mean_s +
                              d_pos_overlap_3gram_mean_s +
                              d_syntactic_dependency_1gram_mean_s +
                              d_syntactic_dependency_2gram_mean_s +
                              d_syntactic_dependency_3gram_mean_s +
                              d_semantic_cosine_mean_s +
                              d_semantic_euclidean_mean_s +
                              (1|WorkerId) + (1|item)
```

```{r}
calc_test_loglik_gaussian <- function(model, test_data, response_var,
                                  allow_new_levels = TRUE) {
  
  mu_pred <- predict(model, newdata = test_data,
                     allow.new.levels = allow_new_levels)
  
  # sd of residuals
  resid_sd <- sigma(model)
  
  observed <- test_data[[response_var]]
  n <- nrow(test_data)
  
  # residual sum of squares
  residual_ss <- sum((test_data$Reaction - mu_pred)^2)
  
  log_lik <- (n / 2) * log(2 * pi) - (n / 2) * log((resid_sd)^2) - (residual_ss / (2 * (resid_sd)^2))
  
  return(
    # sum(dnorm(observed, mean = mu_pred, sd = resid_sd, log = TRUE))
    log_lik
    )
}
```


```{r}
model_cross_val <- function(model_formula, df, group_var, num_folds=10){
  
  # sample folds grouped by WorkerId
  groups <- unique(df[[group_var]])
  fold_map <- setNames(sample(rep(1:num_folds, length.out = length(groups))), groups)
  df$.fold <- fold_map[as.character(df[[group_var]])]
  
  cv_results <- vector("list", num_folds)
  fixef_results <- vector("list", num_folds)
  
  response_var <- all.vars(model_formula)[1]
  
  # create a dummy null model for r2 calculation
  null_formula <- make_null_formula(model_formula)
  
  for (i in 1:num_folds) {
    train_data <- df %>% filter(.fold != i)
    test_data  <- df %>% filter(.fold == i)
    
    null_fit <- lmer(null_formula, data = train_data)
    model_fit <- lmer(model_formula, data = train_data)
    model_log_lik <- calc_test_loglik_gaussian(model_fit, test_data, response_var)
    model_rmse <- calc_test_rmse(model_fit, test_data, response_var)
    model_r2 <- r2(model_fit, null_model = null_fit)

    cv_results[[i]] <- data.frame(
      fold = i, log_lik = model_log_lik, rmse = model_rmse, 
      r2_marginal = as.numeric(model_r2$R2_marginal),
      r2_conditional = as.numeric(model_r2$R2_conditional)
    )
    
    fixef_results[[i]] <- tidy(model_fit, effects = "fixed", conf.int = TRUE) %>%
      transmute(fold = i, term, estimate, std.error, statistic, p.value, 
                conf.low, conf.high)
  }
  
  cv_summary <- do.call(rbind, cv_results)
  fixef_summary <- do.call(rbind, fixef_results)
  
  return(
    list(cv_summary = cv_summary, fixef_summary = fixef_summary)
  )
}
```

```{r}
i <- 1
natstories_fixef_results <- vector("list")
natstories_cv_results <- vector("list")

for (m in unique(natstories_fit_df$model)) {
  for (c in unique(natstories_fit_df$checkpoint)) {
    # print(glue("{m}, {c}"))
    results <- model_cross_val(
      iv_formula,
      natstories_fit_df %>% filter(model == m, checkpoint == c),
      "WorkerId",
      num_folds = 5
    )    
    natstories_fixef_results[[i]] <- results$fixef_summary %>% mutate(model = m, checkpoint = c)
    natstories_cv_results[[i]] <- results$cv_summary %>% mutate(model = m, checkpoint = c)
    
    print(glue("Finished {m}-{c}"))
    
    i <- i + 1
    
  }
}
```
```{r}
natstories_fixef_results0 <- bind_rows(natstories_fixef_results) %>%
  dplyr::select(!c("fold")) %>%
  # group_by(term, model, checkpoint) %>%
  summarise(
    across(everything(),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd = ~sd(.x, na.rm = TRUE),
                serr = ~sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x)))),
            .names = "{.col}_{.fn}"),
    .by = c(term, model, checkpoint)
  ) %>%
  mutate(
    significance_levels = cut(
      p.value_mean,
      breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
      labels = c("***", "**", "*", ".", "ns")
    )
  )

natstories_cv_results0 <- bind_rows(natstories_cv_results) %>%
  dplyr::select(!c("fold")) %>%
  # group_by(model, checkpoint) %>%
  summarise(
    across(everything(),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd = ~sd(.x, na.rm = TRUE),
                serr = ~sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x)))),
            .names = "{.col}_{.fn}"),
    .by = c(model, checkpoint)
  )
```

```{r}
write.csv(natstories_fixef_results0, "results/natstories_fixefs.csv")
write.csv(natstories_cv_results0, "results/natstories_cv.csv")
```

```{r}
natstories_cv_results0 %>%
  ggplot(aes(model, r2_marginal_mean)) +
  geom_col() +
  geom_errorbar(aes(ymin = r2_marginal_mean - r2_marginal_serr, ymax = r2_marginal_mean + r2_marginal_serr), width = 0.2, color = "grey70") +
  facet_grid(~checkpoint) +
  theme_bw()
```

```{r}
natstories_fixef_results0 %>%
  filter(term != "(Intercept)", model == "pythia-12b") %>%
  ggplot(aes(term, estimate_mean, fill = checkpoint)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = estimate_mean - std.error_mean, ymax = estimate_mean + std.error_mean),
                position = position_dodge(width = 0.8),
                width = 0.2, color = "grey70") +
  geom_text(aes(y = 1., label=significance_levels)) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                                "#0072B2", "#D55E00", "#CC79A7", "#999999")) +
  labs(
    y = "Estimate",
    x = "IV value",
    fill = "Checkpoint"
  ) +
  theme_bw() +
  # facet_grid(~c)
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

